Well-quasi-orderings on word languages

### 2025-10-10 src/infixes.tex
lemma 9 polishing

\AP A subset $I \subseteq X$ is \intro(subset){directed} if, for every $x,y \in
I$, there exists $z \in I$ such that $x \leq z$ and $y \leq z$. Given a
\kl{well-quasi-order} $(X, \leq)$, one can always decompose $X$ into a finite
union of \intro{order ideals}, that is, non-empty sets $I \subseteq X$ that are
\kl{downwards closed} and \kl(subset){directed} for the relation $\leq$. In our
case, a \kl{well-quasi-ordered} \kl{order ideal} for the \kl{infix relation} is
the set of finite \kl{infixes} of a finite or bi-infinte word $w \in
\Sigma^\Nat \cup \Sigma^\Rel $ (\cref{bi-infinite:lem}). Therefore, we will
first study languages that are described as the set of finite \kl{infixes} of
some (bi-)infinite word (\cref{ultimately-uniformly-recurrent:lem}). To that
end, let us introduce the notation $\intro*\infset{w}$ for the set of finite
\kl{infixes} of a (possibly infinite or bi-infinite) word $w$.

\begin{lemma}
    \label{bi-infinite:lem}
    Let $L \subseteq \Sigma^*$ be an \kl{order ideal} that is \kl{well-quasi-ordered} 
    for the 
    \kl{infix relation}. Then $L$ is the set of finite \kl{infixes}
    of a finite, infinite or bi-infinite word $w$.
\end{lemma}
\begin{proof}
    Let us assume that $L$ is infinite, the case when it is finite 
    is similar, but leads to a finite word.

    Because the alphabet $\Sigma$ is finite, we can enumerate the words of $L$
    as $\seqof{w_i}[i \in \Nat]$. Now, let us build the sequence $\seqof{u_i}[i \in \Nat]$ by induction
    as follows: $u_0 = w_0$, and $u_{i+1}$ is a word that contains $u_i$ and
    $w_i$, which exists in $L$ because $L$ is \kl(subset){directed}. Since $L$ is
    \kl{well-quasi-ordered}, one can extract an infinite set of indices $I
    \subseteq \Nat$ such that $u_i \infleq u_{j}$ for all $i \leq j \in I$.

>>>>

\AP A subset $I \subseteq X$ is \intro(subset){directed} if, for every $x,y \in
I$, there exists $z \in I$ such that $x \leq z$ and $y \leq z$. Given a
\kl{well-quasi-order} $(X, \leq)$, one can always decompose $X$ into a finite
union of \intro{order ideals}, that is, non-empty sets $I \subseteq X$ that are
\kl{downwards closed} and \kl(subset){directed} for the relation $\leq$. In our
case, a \kl{well-quasi-ordered} \kl{order ideal} for the \kl{infix relation} is
the set of finite \kl{infixes} of a finite \todo{$\Sigma^\Nat$ is an infinite word}
or bi-infinte word $w \in
\Sigma^\Nat \cup \Sigma^\Rel $ (\cref{bi-infinite:lem}). Therefore, we will
first study languages that are described as the set of finite \kl{infixes} of
some (bi-)infinite word (\cref{ultimately-uniformly-recurrent:lem}). To that
end, let us introduce the notation $\intro*\infset{w}$ for the set of finite
\kl{infixes} of a (possibly infinite or bi-infinite) word $w$.

\begin{lemma}
    \label{bi-infinite:lem}
    Let $L \subseteq \Sigma^*$ be an \kl{order ideal} 
    for a \kl{well-quasi-ordered} 
    \kl{infix relation}. Then $L$ is the set of finite \kl{infixes}
    of a finite, infinite or bi-infinite word $w$.
\end{lemma}
\begin{proof}
    Let us assume that $L$ is infinite. The case when it is finite 
    is similar, but will result in a finite word.

    Because the alphabet $\Sigma$ is finite, we can enumerate the words of $L$
    as $\seqof{w_i}[i \in \Nat]$. From $\seqof{w_i}[i \in \Nat]$, we construct a sequence $\seqof{u_i}[i \in \Nat]$ by induction
    as follows: $u_0 = w_0$, and $u_{i+1}$ is a word that contains $u_i$ and
    $w_i$, which exists in $L$ because $L$ is \kl(subset){directed}. Since $L$ is
    \kl{well-quasi-ordered}, one can extract an infinite set of indices $I
    \subseteq \Nat$ such that $u_i \infleq u_{j}$ for all $i \leq j \in I$.
